\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[numbers]{natbib}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  framesep=4pt,
  backgroundcolor=\color{gray!5},
  showstringspaces=false,
}

\title{Critical error rate in practice:\\context-aware variant significance for nanopore amplicon pipelines}
\author{Josh Walker}
\date{}

\begin{document}
\maketitle

\begin{abstract}
The critical error rate (CER) framework~\cite{cer_paper} provides a per-variant
significance statistic $p^*$ that is interpretable against known platform error
rates. The framework as originally posed assumes a single context-free error
rate, which is appropriate for non-homopolymer substitutions in nanopore data
but breaks down for homopolymer (HP) length variants, where empirical error
rates vary by a factor of two or more across HP length and base context~\cite{hp_paper}.
We present a context-aware extension of the framework, retaining the single
binomial equation but replacing the uniform error rate $p$ with a
context-specific $q_\text{ctx}$ derived from empirical measurement. To preserve
cross-variant comparability under context-aware $q_\text{ctx}$, we introduce
the \emph{CER factor}, $p^*/q_\text{ctx}$, as the primary derived metric. The
factor is dimensionless, comparable across variants and specimens, degrades
linearly under $q_\text{ctx}$ misspecification, and re-scales cleanly when the
empirical error table is updated. We describe the implementation in the
speconsense pipeline, including N-denominator choice for post-phasing
validation, identity-group construction, pairwise CER comparison, and a
metadata reproducibility design that allows stored output to be re-evaluated
against revised assumptions without re-running the pipeline.
\end{abstract}

\section{Introduction}

The critical error rate (CER) framework, introduced in~\cite{cer_paper}, asks
the inverse of the conventional p-value question: rather than testing whether
an observed variant is significant under an assumed error rate, it derives the
per-position error rate $p^*$ at which the observed pattern would become
plausible as artifact. A practitioner then compares $p^*$ against their
platform's known error rate; well-supported variants produce $p^*$ values
several times the platform rate, providing a transparent margin for judgment.

The framework's strength is that $p^*$ has the same units as the platform error
rate the reader already knows, decoupling the analysis from the moving target
of basecaller and chemistry improvements. A $p^*$ of $6.8\%$ is robust against
any reasonable nanopore configuration without committing to a specific assumed
rate at output time.

\subsection{The context-dependence problem}

The framework as originally formulated assumes a single per-position error rate
$p$ characterizes the platform. For non-homopolymer substitutions in current
nanopore chemistry, this is approximately true: empirical post-clustering rates
are tightly clustered around 0.6\%--1.0\% across position contexts~\cite{hp_paper}.
The framework's $p^* = 6.8\%$ headline against an assumed rate of $1.5\%$ is
robust to this spread.

For homopolymer-length variants, the assumption breaks down. Homopolymer (HP)
length errors are the dominant systematic error mode in nanopore
sequencing~\cite{delahaye2021}, and their empirical rates depend strongly on
context:

\begin{itemize}
\item HP length error rate increases monotonically from $\sim$0.7\% per position
at $L = 1$ to $\sim$1.5\% at $L = 5$, and steeply thereafter~\cite{hp_paper}.
\item G/C homopolymers consistently produce higher error rates than A/T at
matched length, with the gap widening at longer lengths.
\item Deletion-biased errors dominate insertion-biased errors at a $3$--$5\!:\!1$
ratio across moderate HP lengths.
\item Per-run and per-basecaller variation contribute additional spread of
$\sim$0.6$\times$ to $2.2\times$ on top of the context dependence.
\end{itemize}

A single assumed error rate cannot serve all of these contexts simultaneously
without either over-suppressing variants in low-error contexts (e.g., $L = 1$ HP
positions) or under-suppressing in high-error contexts (e.g., $L = 5$ G runs).
The current speconsense default of $1.5\%$ assumed error rate is calibrated to
cover the non-HP substitution regime; applied uniformly to HP variants, it
loses 30--50\% of the available signal.

\subsection{Contributions}

This paper extends the CER framework to context-dependent error rates without
abandoning its core appeal:

\begin{enumerate}
\item A \textbf{unified equation} that retains the single binomial structure
of~\cite{cer_paper} but replaces $L$ with a count of independent test sites
$n_\text{sites}$ and replaces $q$ with a context-specific $q_\text{ctx}$.
For multi-position variants ($K > 1$), the joint per-read error probability
becomes $\prod_i q_{\text{ctx},i}$ instead of $q^K$, accommodating mixed
contexts naturally.

\item A \textbf{practical decision metric}, the \emph{CER factor}, defined as
$p^* / q_\text{ctx}$. The factor is dimensionless and context-comparable by
construction, recovering the cross-variant comparability that raw $p^*$ loses
under context-aware $q_\text{ctx}$. We compare it against alternative metrics
(raw $p^*$, p-value) and recommend it as the primary annotation.

\item Implementation choices in the \textbf{speconsense pipeline} that follow
from the framework: how $N$ is chosen for post-phasing validation, how
identity groups define the comparison set for pairwise CER, why phasing
deliberately operates without statistical gating, how $q_\text{ctx}$ is
parameterized, and how HP detection is reordered relative to homopolymer
normalization.

\item A \textbf{metadata reproducibility design} that stores enough information
in each specimen's output to recompute $p^*$, $p$-value, and CER factor under
revised assumptions, without re-running the upstream pipeline.
\end{enumerate}

This paper assumes the reader is familiar with the theoretical CER framework
of~\cite{cer_paper} and the empirical HP error characterization
of~\cite{hp_paper}. Section references to these companion documents are made
where useful but the core results are not re-derived.

\section{Unified equation}
\label{sec:unified}

We retain the binomial structure of the CER paper~\cite{cer_paper}. The
generalization replaces the amplicon length $L$ with a count of independent test
sites $n_\text{sites}$ and replaces the single per-position error probability
$q$ with a context-specific $q_\text{ctx}$.

\subsection{Single-position variants ($K = 1$)}

For a candidate variant supported by $M$ reads at a single position with context
$c$ in a specimen of $N$ total reads,
\begin{equation}
\label{eq:unified_k1}
n_\text{sites} \cdot P\bigl(\mathrm{Binom}(N,\, q_{\text{ctx},c}) \geq M\bigr) = T,
\end{equation}
where $T$ is either an input significance level $\alpha$ (when solving for
$p^*$) or an output p-value (when $q_{\text{ctx},c}$ is fixed and the equation
is evaluated). This reduces to the formulation in~\cite{cer_paper} when
$q_{\text{ctx},c} = q$ is uniform across positions and
$n_\text{sites} = L$.

\subsection{Multi-position variants ($K > 1$)}

When a variant is distinguished from its comparison cluster at $K > 1$
positions, the joint per-read probability of erroring at all $K$ positions under
the null is the product of per-position rates:
\begin{equation}
\label{eq:unified_kn}
\binom{n_\text{sites}}{K} \cdot P\Bigl(\mathrm{Binom}\bigl(N,\, \textstyle\prod_{i=1}^{K} q_{\text{ctx},i}\bigr) \geq M\Bigr) = T.
\end{equation}
This generalization is essential for HP applications because mixed-context
$K > 1$ variants are common: a multi-position variant may be distinguished at
one HP position and one non-HP position simultaneously. Equation~\ref{eq:unified_kn}
reduces to $\binom{L}{K} \cdot P(\mathrm{Binom}(N, q^K) \geq M) = T$ when all
$q_{\text{ctx},i}$ are equal.

\subsection{Bonferroni site counting}
\label{sec:nsites}

The site count $n_\text{sites}$ is the number of independent positions tested.
Three definitions are operationally distinct:

\begin{description}
\item[Per-base count.] $n_\text{sites} = L$, the amplicon length in bases.
This is the formulation in~\cite{cer_paper}; appropriate when every position
is tested independently.

\item[Per-site count with HP aggregation.] $n_\text{sites} = (\text{non-HP positions}) + (\text{HP runs of } L \geq 2)$.
Equivalently, $n_\text{sites}$ is the length of the \emph{homopolymer-compressed}
consensus --- the sequence obtained by collapsing each maximal run of identical
bases to a single base (so \texttt{CAAAGGGG} has three sites: one per distinct
run). A length-5 HP run is one site for length-variant testing, not five. For
ITS amplicons of $\sim$700~bp, this aggregation reduces $n_\text{sites}$
modestly (typically by $\sim$15--25\%) because most HP runs are short.

\item[Per-site count with alternative-base correction.] Each non-HP position
admits 3 alternative bases; each $L \geq 2$ HP run admits multiple alternative
lengths. A strict Bonferroni would multiply $n_\text{sites}$ by the number of
plausible alternatives per site. Following~\cite{cer_paper}, we treat the $q
\to q/3$ uniform error split as implicit per-position correction over alternative
bases for non-HP positions, and use empirical per-transition rates for HP
positions (which are already specific to one alternative).
\end{description}

We adopt the per-site count with HP aggregation as the default. The choice
affects $p^*$ only logarithmically (a 25\% reduction in $n_\text{sites}$
shifts $p^*$ by less than 5\%).

\section{The CER factor}
\label{sec:factor}

\subsection{Definition}

The \emph{CER factor} is the dimensionless ratio
\begin{equation}
\label{eq:factor}
\mathrm{CER\ factor} = \frac{p^*}{q_\text{ctx}}.
\end{equation}
It answers the question: \emph{by what factor would the platform error rate at
this variant's context need to be inflated to make the observation plausible as
artifact?} A factor of 4 means quadruple the empirical error rate would be
needed; a factor of 1 corresponds exactly to the platform rate.

\subsection{Comparison with alternative metrics}

Three derived metrics summarize the same underlying CER surface:

\begin{description}
\item[Raw $p^*$.] The original CER paper's headline metric. Interpretable as
``the per-position error rate that would explain this variant.'' Comparable
across variants only when $q_\text{ctx}$ is approximately uniform.

\item[p-value.] The conventional statistic, computed by fixing $q_\text{ctx}$
and evaluating Equation~\ref{eq:unified_k1} for $T$. Universal 0--1 scale,
composable for FDR and meta-analysis, exponentially sensitive to
$q_\text{ctx}$ misspecification.

\item[CER factor.] Defined by Equation~\ref{eq:factor}. Dimensionless,
context-independent on the output side, linearly sensitive to $q_\text{ctx}$
misspecification, re-scalable when $q_\text{ctx}$ updates.
\end{description}

Table~\ref{tab:metric_comparison} summarizes the trade-offs.

\begin{table}[htbp]
\centering
\caption{Trade-off summary for derived CER metrics. ``$\checkmark$'' indicates
the metric is preferred on that dimension.}
\label{tab:metric_comparison}
\small
\begin{tabular}{lccc}
\toprule
Property & Raw $p^*$ & p-value & CER factor \\
\midrule
Context-comparable across variants                  &              &  $\checkmark$ & $\checkmark$ \\
Linear sensitivity to $q_\text{ctx}$ error          &              &              & $\checkmark$ \\
Composable for FDR / meta-analysis                  &              & $\checkmark$ &              \\
Principled default decision threshold               &              & $\checkmark$ &              \\
Reinterpretable against future chemistry            & $\checkmark$ &              & $\checkmark$ \\
Familiar to general statistical audience            &              & $\checkmark$ &              \\
Bounded dynamic range at strong end                 &              &              & $\checkmark$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Why the factor is the right primary metric for amplicon pipelines}

For speconsense's primary use case --- amplicon barcoding where consensus
sequences are deposited into taxonomic reference databases --- the CER factor's
properties align with what practitioners need:

\begin{enumerate}
\item \textbf{Cross-variant comparability is the dominant requirement.}
Validators rank variants by confidence and apply a threshold. The factor
provides a single comparable scale across HP and non-HP contexts.

\item \textbf{Linear sensitivity to $q_\text{ctx}$ is honest.} Empirical
error rates vary $\sim$0.6$\times$ between sequencing runs on the same
basecaller~\cite{hp_paper}. P-values would swing by orders of magnitude under
this drift; the factor moves linearly, which more accurately reflects our
uncertainty about the true error rate at any given site.

\item \textbf{Re-scaling under updated chemistry is direct.} If a future
$q_\text{ctx}$ table revises the rate at a context by a factor $r$, stored
factors at that context can be rescaled by $1/r$ without recomputing from raw
$N$ and $M$. P-values must be recomputed from raw inputs; raw $p^*$ does not
require rescaling but loses cross-context comparability.

\item \textbf{The robustness interpretation survives.} A factor of 4 is
``quadruple the empirical floor would be needed to explain this as artifact''
--- a direct generalization of the raw $p^*$ interpretation, retaining the
human-interpretable units that motivated CER in the first place.
\end{enumerate}

\subsection{Default decision threshold}

We propose a default factor threshold of $\geq 2$ for accepting a variant: the
observed pattern requires at least double the empirical context-specific error
rate to be explained as artifact. The threshold is conservative under typical
$q_\text{ctx}$ uncertainty (run-to-run variation, basecaller drift) and
recovers the boundary behavior of the existing default $p^* \geq 1.5\%$
against a non-HP $q_\text{ctx} \approx 0.7\%$ (factor $\approx$ 2.1).

The threshold is a tunable: stricter calls (factor $\geq 3$) are appropriate
for novel reference deposits; permissive calls (factor $\geq 1$) are
appropriate for diversity surveys where sensitivity matters more than
specificity. Section~\ref{sec:reproducibility} describes how the factor can be
recomputed at any threshold from stored metadata without re-running the
pipeline.

In the initial deployment the production default is factor $\geq 1.0$, not
the factor $\geq 2$ analyzed above. This is a pragmatic interim choice: the
factor $= 1$ boundary is the straightforward ``observation explainable by the
empirical rate alone'' cut, easy to reason about while the pipeline accumulates
validation data. The threshold is expected to be tuned upward (likely toward
factor $\geq 2$ for deposit-quality calls) once the empirical distribution of
factor values on real workloads is characterized. Because the factor is recorded
per-variant in the metadata JSON (Section~\ref{sec:reproducibility}), threshold
tuning does not require re-running the upstream pipeline.

\subsection{FASTA annotation scheme}
\label{sec:fasta_annotation}

A single CER field is emitted per consensus header; the forensic detail lives
in the per-specimen metadata JSON (Section~\ref{sec:reproducibility}).

\begin{lstlisting}
>sample-1.v2 size=200 ric=100 rid=95.2 cer_factor=4.2
\end{lstlisting}

The header carries only \texttt{cer\_factor}: a single dimensionless number,
the primary decision metric; higher means more robust. Per-position detail
($p^*$, $K$, context tags, $q_\text{ctx}$ values, reference cluster index)
lives in the metadata JSON.

An earlier draft of this framework emitted a parallel \texttt{cer\_details}
field in the header encoding $p^*$, $K$, and context/$q$ pairs joined by
\texttt{+}. We abandoned this in implementation: at $K > 1$ the field becomes
a multi-token string that crowds the fixed-width \texttt{-all.fasta} header,
inflates line length past comfortable terminal display, and duplicates data
the metadata JSON already carries in structured form. The header is the
triage surface (\emph{does this variant pass?}); the JSON is the forensic
surface (\emph{why does this variant pass?}). Splitting by role keeps the
header scannable and the forensic detail queryable.

\section{Parameterizing $q_\text{ctx}$}
\label{sec:parameterization}

The unified framework demands a context classifier and a $q_\text{ctx}$ table.
This section describes the context taxonomy, the source of empirical values,
and the operational choices for keeping $q_\text{ctx}$ current.

\subsection{Context classes}

Each MSA position falls into one of a small set of context classes. The
classification is deterministic from the consensus sequence at that position
and depends on no upstream call:

\begin{description}
\item[Non-HP substitution.] The consensus base is not part of an HP run of
length $\geq 2$, and the variant differs by a substitution. Empirical
$q_\text{ctx}$ corresponds to the per-position substitution rate; for Dorado
v5.0 on R10.4.1, $q_\text{ctx} \approx 0.0059$~\cite{hp_paper}.

\item[HP-interface substitution.] The consensus base is the first or last base
of an HP run of length $\geq 2$, and the variant is a substitution. Empirical
analysis~\cite{hp_paper}~\S 4 demonstrates that interface substitutions occur
at rates indistinguishable from non-HP substitutions; they are assigned the
non-HP substitution rate.

\item[HP length change at $L$.] The variant differs from the consensus by a
length change in an HP run of reference length $L \in \{2, 3, 4, 5\}$.
Empirical $q_\text{ctx}$ is read from the per-$L$ table
in~\cite{hp_paper}~\S 3.3, pooled across base and direction
(Section~\ref{sec:qctx_pooling}).

\item[HP length change at $L \geq 6$.] Suppressed by blanket normalization;
no $q_\text{ctx}$ assigned and no CER evaluation. This follows the
recommendation of~\cite{hp_paper}~\S 8.1: at $L \geq 6$ accuracy degrades
sharply and variant calling becomes unreliable.

\item[Indel at non-HP position.] Treated as a substitution-class event with
the non-HP rate as a placeholder. Indels at non-HP positions are rare in
amplicon data and do not warrant a separate empirical column at this stage.
\end{description}

The classifier returns a context tag and a $q_\text{ctx}$ value. The context
tag records the base and direction of an HP-length event for output
annotation (e.g., \texttt{hp-l3-A-del1}), even though the $q_\text{ctx}$
lookup uses only the length. For mixed-context $K > 1$ variants, the
per-position tags are
recorded as a list and the joint $q_\text{ctx}$ is the product
(Equation~\ref{eq:unified_kn}).

\subsubsection{Pooling base and direction in HP $q_\text{ctx}$}
\label{sec:qctx_pooling}

The HP length-change class could in principle split along three sub-dimensions:
base $b \in \{A, T, C, G\}$, reference length $L$, and direction/magnitude
$\delta$ (deletion vs. insertion, one-base vs. multi-base). A fully specified
table would require $\sim\!4 \times 4 \times \geq\!2 = 32+$ entries for the
$L \leq 5$ regime alone. The shipped tables collapse $b$ and $\delta$, keeping
only $L$: five HP entries plus two non-HP entries. This is a deliberate
simplification driven by empirical signal-to-noise in~\cite{hp_paper}:

\begin{itemize}
\item \textbf{Base (pooled).} Per-base variation at $L \leq 5$ on Dorado v5.0
is small: maximum over minimum across bases is $1.14$ at $L = 1$ rising to
$1.39$ at $L = 5$~\cite{hp_paper}~\S 7.6. The ``worst-case base'' is not
consistent across lengths (T at $L = 2$--$3$, G at $L = 4$--$5$), so no
stable A/T-vs-C/G split is supported in this regime.

\item \textbf{Direction (pooled).} The deletion-to-insertion ratio is 3--5:1
at moderate lengths~\cite{hp_paper}~\S 3.4, but per-cell counts for
insertions --- especially at $(b, L, \delta = +\!1)$ --- are sparse. Splitting
would produce unstable rates for the less-populated cells.

\item \textbf{Length (kept).} The per-$L$ effect is the largest and most
stable: rates rise monotonically from $0.67\%$ at $L = 1$ to $1.13\%$ at
$L = 5$ on ont98~\cite{hp_paper}~Table~4, a $\sim\!1.7\times$ span resolved
above cross-run noise.
\end{itemize}

The larger error-table uncertainty --- $\sim\!0.6\times$ between sequencing
runs on the same basecaller~\cite{hp_paper}~\S 5.4 --- dominates the
$\sim\!1.4\times$ per-base and $\sim\!3\text{--}5\times$ direction effects
only for the insertion direction; for deletions on pooled data the dominant
axes are length and run-to-run variation, which the current scheme captures.
Finer segmentation by base and direction is straightforward to add (extend
the YAML with \texttt{hp-l$\{L\}$-$\{b\}$-$\{\delta\}$} keys; the lookup
function already accepts arbitrary string keys) if future data shows the
finer cells stabilizing above the run-to-run noise floor. The output context
tag records the base and direction regardless, so downstream re-analysis
against a finer table requires no upstream re-run.

\subsection{Sources of $q_\text{ctx}$ values}

Three operational regimes for $q_\text{ctx}$ are appropriate at different
levels of pipeline maturity:

\begin{description}
\item[Phase 1 (initial deployment): shipped per-basecaller table.] A static
$q_\text{ctx}$ table is bundled with speconsense, parameterized by basecaller
identifier (e.g., \texttt{dorado-v5.0}). Values are drawn from the empirical
analysis in~\cite{hp_paper}, pooled across base and direction as described in
Section~\ref{sec:qctx_pooling}. The table identifier is recorded in metadata
(Section~\ref{sec:reproducibility}). Users select their basecaller via a CLI
flag (\texttt{--error-model dorado-v5.0}) or accept the default.

\item[Phase 2 (mature deployment): per-run estimation.] $q_\text{ctx}$ is
estimated from the current run's MSAs at pipeline startup. The estimator
pools positions across all specimens after applying the
outlier-and-bimodal-detection biological-variant filter
of~\cite{hp_paper}~\S 5.3. The estimated $q_\text{ctx}$ is recorded in
metadata so that downstream interpretation can use the run-specific values
or fall back to the shipped defaults.

\item[Override: user-supplied table.] A flag (\texttt{--qctx-table-file PATH})
accepts a TSV with columns \texttt{(context\_tag, q\_ctx)}. Useful for
research applications that need to characterize variant calls under
synthetic or extreme error assumptions.
\end{description}

The first deployment supports Phase 1 only. Phase 2 is a substantial
implementation effort and is left for future work; it is mentioned here so
that the metadata schema (Section~\ref{sec:reproducibility}) is forward-
compatible.

\subsection{Stability across runs}

The cross-run scaling factor between ont98 and run116 (same basecaller, three
months apart) is $\sim$0.6$\times$ at $L = 1$--5~\cite{hp_paper}~\S 5.4.
This means a shipped Phase 1 table may differ from a specific run's actual
rates by up to a factor of 2. The CER factor's linear sensitivity makes this
a roughly 2$\times$ shift in factor values, which is large but interpretable;
p-values under the same misspecification would shift by orders of magnitude.
Phase 2 per-run estimation eliminates this source of uncertainty.

\section{Implementation in speconsense}
\label{sec:implementation}

This section describes how the unified framework integrates with the existing
speconsense pipeline. The current implementation~\cite{speconsense} performs
recursive MSA-based variant phasing, post-phasing identity grouping, and
pairwise CER validation. The context-aware framework requires changes at each
of these stages.

\subsection{N denominator choice}
\label{sec:n_choice}

CER is computed only at post-phasing pairwise validation
(Section~\ref{sec:phasing_design}); phasing itself uses local frequency and
count thresholds without statistical gating, and any pass/fail decision from
\texttt{cer\_factor} is deferred to summarize
(Section~\ref{sec:annotation_vs_gating}). The denominator $N$ in the binomial
therefore has a single definition at the only point where it enters the
computation.

\paragraph{Identity-group sum.} When validating one cluster against another
within an identity group, $N$ is the sum of read counts within the group, not
the total specimen count. The pairwise comparison's null hypothesis is
``these two clusters represent the same underlying sequence, with the smaller
cluster arising as artifact of the larger''---a question that operates on the
population of reads sharing the same identity-group membership, not on the
broader specimen pool.

In the current implementation, this is computed in
\texttt{\_validate\_identity\_group} as
\texttt{group\_N = sum(len(subclusters[i]['read\_ids']) for i in sorted\_indices)}.

\paragraph{Why the identity-group sum is the right choice.} The artifact
hypothesis under test is ``the candidate cluster is misclassified reads from
its sibling.'' The relevant population is the union of reads that could
plausibly have ended up in either cluster --- the identity group. Using the
total specimen reads as $N$ would inflate the denominator relative to the
hypothesis under test, producing artificially small $p^*$ values (and CER
factors) and over-accepting weak variants. Using only the candidate cluster's
own size as $N$ would conversely understate the denominator and over-reject.
The identity-group sum corresponds exactly to the population over which the
candidate is hypothesized to be a redistribution of reads from validated
siblings.

\paragraph{Why not ``total specimen reads''.} A natural alternative is to use
the total specimen input reads as $N$, on the conservative reasoning that any
read in the specimen could in principle have appeared in this position. We
considered and rejected this. The artifact null is not ``any specimen read
could have erred here'' --- which is what total specimen $N$ models --- but
``among the reads belonging to this identity group, could redistribution have
produced the candidate's count.'' Reads outside the group are part of
unrelated identity groups and do not enter the artifact mechanism for this
candidate.

\subsection{Identity groups}
\label{sec:identity_groups}

An \emph{identity group} is a set of clusters whose pairwise HP-normalized
identity exceeds a threshold for \emph{every} pair in the group. Groups define
the comparison set for pairwise CER: a candidate variant is tested against
members of its own identity group, not against the entire specimen.

\paragraph{Construction.} Groups are formed by complete-linkage hierarchical
clustering (\texttt{scipy.cluster.hierarchy}) on pairwise HP-normalized
adjusted-identity distance, cut at $1 - \texttt{group\_identity}$. The default
threshold is 85\% identity (15\% adjusted-identity distance), tunable via
\texttt{--group-identity}. Under complete linkage, every pair of members in a
group satisfies the identity threshold; under single-linkage union-find, only
a single connecting chain is required.

\paragraph{Why complete linkage over union-find.} We adopted complete linkage
after observing transitive-chain failures on environmental-DNA specimens. eDNA
samples commonly contain taxonomically distinct but related sequences (e.g.,
sibling species amplified together) that would form a single chain under
union-find at any reasonable threshold, collapsing real organismal diversity
into one identity group and producing misleading CER comparisons between
unrelated amplicons. Complete linkage ensures that grouping survives the
``furthest pair'' test: two clusters whose pairwise identity is below threshold
cannot end up in the same group via an intermediate. This matches the
within-specimen identity criterion we want the group to enforce.

\paragraph{Role in CER decisions.} Within each identity group, clusters are
ordered by size. The largest cluster (the \emph{anchor}) auto-passes CER
without comparison. Each smaller cluster is tested pairwise against all
already-validated clusters in the group; the candidate's CER factor is the
\emph{minimum} across all pairwise comparisons, corresponding to the
``nearest plausible artifact source.'' Acceptance requires the minimum
factor to exceed the decision threshold.

\paragraph{Why groups, not any-to-any.} The artifact hypothesis is that a
small cluster represents reads misclassified from a similar cluster. This
hypothesis only makes sense within a similarity group; clusters of distinct
sequences cannot serve as artifact sources for each other. Restricting the
comparison set to identity-group members both encodes this hypothesis and
reduces compute by avoiding $O(n^2)$ comparisons across the specimen.

\subsection{Pairwise CER comparison}
\label{sec:pairwise}

Pairwise CER between candidate cluster $A$ and validated reference cluster $B$
proceeds as follows:

\begin{enumerate}
\item Compute pairwise differences $K = \mathrm{count\_variant\_differences}(A_\text{cons}, B_\text{cons})$,
where each substitution contributes 1 and each contiguous indel event
contributes 1 (regardless of indel length).

\item For each of the $K$ differing positions, classify the context and look
up $q_{\text{ctx},i}$.

\item Compute the joint per-read error probability $\prod_i q_{\text{ctx},i}$.

\item Compute $n_\text{sites}$ over the longer of the two consensuses by
counting maximal base runs (Section~\ref{sec:nsites}): each non-HP position
contributes one site, and each HP run of length $\geq 2$ also contributes one
site regardless of run length.

\item Take $N$ as the identity-group sum (Section~\ref{sec:n_choice}) and $M$
as the candidate cluster's read count.

\item Solve Equation~\ref{eq:unified_kn} for $p^*$ at the configured $\alpha$,
then compute the CER factor.
\end{enumerate}

The candidate's reported CER factor is the minimum over all pairwise
comparisons against validated cluster references. This conservative choice
reflects the worst-case artifact source and ensures that if any nearby
validated cluster could plausibly explain the candidate as artifact, the
candidate is suppressed.

\subsection{Annotation in core, gating in summarize}
\label{sec:annotation_vs_gating}

The speconsense pipeline partitions responsibility across two executables:
\texttt{speconsense} (core) produces consensus sequences with per-variant CER
annotations, and \texttt{speconsense-summarize} consumes the annotated output
and applies user-visible thresholds. CER gating lives exclusively in summarize.

Every non-anchor candidate is pairwise-compared against all larger peers in
its identity group, annotated with a \texttt{cer\_factor} (per-position
multiplicative inflation; worst-case across peers), and flows through to the
FASTA output. Records with \texttt{cer\_factor} below the user-configured
\texttt{--min-cer-factor} are routed by summarize to a separate \texttt{.ns}
(non-significant) output file, mirroring the primary output layout; records
with \texttt{cer\_factor} undefined (anchors, comparisons that failed to
converge) always pass.

This separation is deliberate. CER gating answers a question with a tunable
threshold: which variants should a reader \emph{see} as primary calls? That
threshold depends on the reader's application (reference deposit vs.\
diversity survey), their platform's current error characterization, and the
significance level chosen for the run. Baking the gate into core would force
re-running clustering and consensus generation to retune the threshold.
Keeping annotation in core and gating in summarize makes threshold retuning a
seconds-scale rerun of summarize on the same FASTA, and preserves the
sub-threshold clusters as inspection artifacts rather than destroying them.

\subsection{Phasing without statistical gating}
\label{sec:phasing_design}

Recursive MSA-based phasing in speconsense produces candidate clusters using
only local frequency and count thresholds (\texttt{--min-variant-frequency},
\texttt{--min-variant-count}); CER is not evaluated during phasing. This is
a deliberate design choice, not an oversight, and we intend to keep it that
way under the context-aware framework.

\paragraph{Phasing as candidate generation.} The role of phasing is to
\emph{partition} reads into candidate haplotype clusters based on positional
allele patterns. Phasing answers the structural question ``which reads
co-vary at which positions,'' producing a set of candidate clusters whose
existence is a hypothesis to be tested downstream. The role of CER is to
\emph{evaluate} those candidates against the artifact null. Conflating
generation and evaluation forces every phasing decision through the CER
machinery, which is both expensive and structurally inappropriate: at the
moment a phasing split is being considered, the candidate cluster has not
yet had its consensus generated, so its $K$ relative to siblings is unknown
and the appropriate $q_\text{ctx}$ values cannot be computed.

\paragraph{Why post-phasing CER is preferred.} Three reasons:

\begin{enumerate}
\item \textbf{Mixed-context $K > 1$ requires consensus alignment.} The
unified equation requires per-position $q_{\text{ctx},i}$ values for each of
the $K$ positions distinguishing the candidate from its sibling. These are
known only after both consensuses are generated and aligned, which happens
post-phasing. A phasing-time gate would be limited to $K = 1$ evaluation
against an arbitrary reference, losing the substantial sensitivity gains of
multi-position evaluation (Section~\ref{sec:unified}).

\item \textbf{Phasing's local thresholds are fast and effective pre-filters.}
The existing \texttt{--min-variant-count 5} and
\texttt{--min-variant-frequency 0.10} thresholds eliminate the
overwhelmingly majority of trivial splits (a single-read singleton at a
random position) without any statistical machinery. CER's contribution is
discriminating among the remaining candidates, all of which have at least
modest local support. Running CER at every phasing step would be largely
wasted work.

\item \textbf{Identity-group $N$ depends on phasing's output.} The
identity-group sum used as $N$ in pairwise CER (Section~\ref{sec:n_choice})
is determined by the set of clusters phasing produces and how they group by
similarity. The hypothesis a phasing-time gate would test (``is this
sub-split significant against \emph{some} pool'') has no well-defined $N$
until the full set of clusters is known.

\item \textbf{Cleaner primary consensuses via noise excision.} Indiscriminate
phasing followed by post-validation rejection lets non-significant variants
be cleanly identified as their own \emph{ns} (not significant) clusters
rather than being absorbed back into a parent and manifesting as IUPAC
ambiguity codes. With a phasing-time gate, reads that would have formed a
non-significant split remain in their parent cluster, contributing to
ambiguity at the affected positions and dirtying the parent's consensus. The
no-gate design separates ``structurally a candidate variant'' from
``statistically meaningful,'' so a primary consensus accumulates ambiguity
only from positions where reads genuinely could not be phased apart at
all---not from positions where phasing succeeded but the resulting cluster
was too weakly supported to report. The non-significant reads remain visible
in the output as ns clusters, available for inspection without polluting the
significant consensuses.
\end{enumerate}

\paragraph{What about over-splitting?} A reasonable concern is that
indiscriminate phasing produces clusters that should never have been split,
inflating downstream compute and adding noise. In practice, the local
frequency and count thresholds do most of this work: a 10\% / 5-read floor
prevents single-read splits and cleanly catches obvious noise. Post-phasing
identity-group construction (Section~\ref{sec:identity_groups}) collapses
near-identical splits via union-find, and pairwise CER then suppresses
weakly-supported candidates. The pipeline's three-stage structure
(local-threshold phasing $\to$ similarity-based grouping $\to$ statistical
validation) cleanly separates concerns and matches each stage to the
question it is structurally suited to answer.

\subsection{HP detection and normalization}
\label{sec:hp_normalization}

The pipeline treats HP-length variation at three distinct stages with
deliberately different normalization defaults. Taken together, these stages
admit HP length variants at $L \leq 5$ into CER evaluation with an appropriate
$q_\text{ctx}$ while suppressing the noisier $L \geq 6$ regime, as motivated
by~\cite{hp_paper}~\S 8.1.

\paragraph{Stage 1 --- initial clustering identity.} Initial read clustering
(both the MCL K-NN graph and the greedy algorithm) uses \emph{raw} edlib
identity, $1 - \mathrm{editDistance}/\max(|s_1|,|s_2|)$, computed in
\texttt{SpecimenClusterer.calculate\_similarity}. No homopolymer normalization
is applied at this stage. The threshold \texttt{--min-identity} (default
$0.90$) filters raw pairs. HP length variants at $L \leq 5$ therefore do not
drive separation at this stage for the typical amplicon: a single-base HP-
length disagreement on a $\sim\!700$~bp ITS read contributes $\sim\!0.15\%$ to
raw distance, well inside the $10\%$ tolerance of the default threshold, so
reads with HP length polymorphism pool into the same initial cluster in
practice. This is a threshold-driven outcome, not an explicit HP
normalization; sequences with accumulated error that crosses the threshold
can still separate.

\paragraph{Stage 2 --- phasing variant detection.} MSA-based variant
detection (\texttt{is\_variant\_position\_with\_composition} in
\texttt{speconsense.msa}) subtracts the HP-extension contribution from the
raw base-composition counts before thresholding, but does so only for HP runs
whose length meets the \texttt{min\_hp\_length} threshold of the HP-context
builder (default $6$). At $L \leq 5$ the HP-extension composition is empty,
so the subtraction is a no-op and HP length variants are surfaced as candidate
phasing positions. At $L \geq 6$ the subtraction zeros the variant count and
the position is suppressed. Phasing therefore \emph{does} split an initial
cluster when its reads disagree at an $L \leq 5$ HP-length column, producing
separate consensus clusters that flow downstream.

\paragraph{Stage 3 --- post-phasing HP-equivalence merging.} Clusters whose
consensuses differ only in HP lengths are candidates for merging via
\texttt{are\_homopolymer\_equivalent}, again with \texttt{min\_hp\_length} of
$6$. Two clusters differing only at an $L \leq 5$ HP position are not merged
and are carried forward as distinct variants; differences at $L \geq 6$ are
treated as HP-equivalent and collapsed.

The three stages together enforce the following regime for HP length
variation. At $L \leq 5$: Stage 1 pools reads through threshold tolerance
(raw single-base indels are below the $10\%$ distance budget), Stage 2
surfaces the column as a phasing candidate, and Stage 3 leaves the resulting
split intact. At $L \geq 6$: Stage 1 still pools most reads through threshold
tolerance, Stage 2 suppresses the variant column entirely, and Stage 3
collapses any residual split via HP-equivalence merging. The
\texttt{min\_hp\_length} defaults at Stages 2 and 3 are intentionally aligned
to $6$: lowering it at Stage 2 would let high-error long runs drive spurious
splits, and lowering it at Stage 3 would erase the short-HP variants that
Stage 2 just surfaced.

\paragraph{Downstream handling.} Surviving $L \leq 5$ HP-length variants
reach pairwise CER evaluation as $K = 1$ candidates whose single differing
position carries an \texttt{hp-l\{L\}-\{base\}-\{dir\}\{count\}} context tag
(e.g., \texttt{hp-l3-A-del1}; Section~\ref{sec:parameterization}). The
context classifier's \texttt{hp\_min\_length} is set to $2$ so that any run
of two or more bases contributes an HP tag to pairwise comparisons; the
$L \leq 5$ vs $L \geq 6$ question has already been settled by Stages~2 and
3, so the classifier does not need to re-enforce it.

\paragraph{Ambiguity handling.} The pipeline does not currently absorb
sub-threshold HP length variants into the parent consensus as explicit
HP-ambiguity annotations; variants that fail CER in summarize are routed to
the \texttt{.ns} output (Section~\ref{sec:annotation_vs_gating}) and the
primary consensus retains its own modal HP length. Encoding HP length
variation as a structured header annotation (e.g., \texttt{poly\_A[5..6]})
for downstream inspection remains an open design question.

\section{Metadata reproducibility}
\label{sec:reproducibility}

The CER computation depends on several inputs --- $N$, $M$, $K$,
$n_\text{sites}$, $q_\text{ctx}$, $\alpha$ --- and a stable mapping from
position to context. As empirical $q_\text{ctx}$ values are revised (new
basecaller versions, per-run estimation, improved error characterization),
practitioners should be able to recompute CER factor and p-value from
\emph{stored} metadata, without re-running the pipeline. This section
specifies the metadata schema that supports this.

\subsection{Specimen metadata JSON}

Each specimen produces a metadata JSON file at the standard output location.
The current schema~\cite{speconsense} contains parameters, input information,
and primer settings; we extend it with CER reproduction data:

\begin{lstlisting}
{
  "schema_version": "2.0",
  "version": "0.7.7",
  "sample_name": "ONT08.75-C10--iNat262724197",
  "timestamp": "2026-04-20T18:09:14.624236",
  "parameters": {
    "algorithm": "graph",
    "min_identity": 0.9,
    "min_size": 3,
    "significance_level": 1e-5,
    "group_identity": 0.85,
    "error_model": "dorado-v5.0",
    ...
  },
  "total_input_reads": 1000,
  "identity_groups": [
    {
      "group_id": "g1",
      "group_N": 974,
      "members": ["1.v1", "1.v2", "1.v3", ...]
    }
  ],
  "variants": [
    {
      "cluster_id": "1.v1",
      "identity_group": "g1",
      "group_rank": 1,
      "variant_rank": 1,
      "M": 462, "N": 974,
      "K": null, "n_sites": null, "context_tags": null,
      "q_ctx_per_position": null,
      "compared_against_idx": null,
      "cer_factor": null, "cer_pstar": null,
      "err_factor": 0.303,
      "err_factor_obs_sum": 3.13,
      "err_factor_exp_sum": 10.33,
      "err_factor_cols": 781
    },
    {
      "cluster_id": "1.v3",
      "identity_group": "g1",
      "group_rank": 1, "variant_rank": 3,
      "M": 49, "N": 974, "K": 1, "n_sites": 561,
      "context_tags": ["hp-l3-A-del1"],
      "q_ctx_per_position": [0.0097],
      "compared_against_idx": 0,
      "cer_factor": 2.085,
      "cer_pstar": 0.0202,
      "err_factor": 0.453,
      "err_factor_obs_sum": 4.67,
      "err_factor_exp_sum": 10.32,
      "err_factor_cols": 780
    }
  ]
}
\end{lstlisting}

Cluster identifiers use the \texttt{\{group\_rank\}.v\{variant\_rank\}} form
assigned by core (e.g., \texttt{1.v3} is the third-largest variant within the
largest identity group). The anchor of each identity group is the entry with
\texttt{compared\_against\_idx = null}; all CER fields for anchors are null by
construction. Non-anchor variants carry the integer index (into the variant
list) of the reference cluster against which \texttt{cer\_factor} was computed.
Fields \texttt{err\_factor}, \texttt{err\_factor\_obs\_sum},
\texttt{err\_factor\_exp\_sum}, and \texttt{err\_factor\_cols} record the
cluster homogeneity statistic described in Section~\ref{sec:denoising} along
with its raw components; downstream tools can recompute the factor under a
revised $q_\text{ctx}$ table without re-running the pipeline.

Derived scalars that are cheap to recompute from the stored inputs ---
$q_\text{ctx,joint} = \prod_i q_{\text{ctx},i}$ and the p-value at the stored
$q_\text{ctx}$ --- are not persisted.

\subsection{Recomputation workflow}

Given a stored metadata JSON and a revised $q_\text{ctx}$ table or significance
level, recomputation proceeds:

\begin{enumerate}
\item For each variant, look up the new $q_{\text{ctx},i}$ values for the
recorded \texttt{context\_tags}.
\item Compute the new joint $q_\text{ctx,joint}^\text{new} = \prod_i q_{\text{ctx},i}^\text{new}$.
\item Recompute $p^*$ from $(N, M, n_\text{sites}, K, \alpha^\text{new})$.
\item The CER factor becomes $p^* / q_\text{ctx,joint}^\text{new}$ (or, for
cross-variant reporting on a per-position scale, $(p^*)^{1/K} / \bar{q}_\text{ctx}^\text{new}$
where $\bar{q}_\text{ctx}^\text{new}$ is the geometric mean of the new
per-position rates; the implementation currently reports the joint form).
\item If a p-value is needed, compute
$n_\text{sites} \cdot P(\mathrm{Binom}(N, q_\text{ctx,joint}^\text{new}) \geq M)$.
\end{enumerate}

The original $N$, $M$, $K$, context tags, and consensus-derived
$n_\text{sites}$ are immutable properties of the sequencing data; no upstream
pipeline run is needed to revise the significance metric.

\subsection{Schema versioning}

The \texttt{schema\_version} field allows downstream tools to detect missing
fields and degrade gracefully. Version 1.0 is the pre-context-aware schema;
version 2.0 introduces the CER fields in this paper. Tools reading version 1.0
output can fall back to the original raw-$p^*$ interpretation; tools reading
version 2.0 output have full reproducibility.

\subsection{Why this is worth the bytes}

The metadata addition adds $\sim$200--500 bytes per variant. For a typical
ITS amplicon study with 1000 specimens and an average of 3 variants each, the
total overhead is on the order of 1--2~MB across the run. This is negligible
relative to the FASTA and FASTQ output. The benefit --- recomputation under
revised assumptions without re-running --- is substantial: as basecaller
versions evolve and empirical error tables improve, every stored output
remains scientifically valid without recomputation cost.

\section{Cluster denoising precedes CER}
\label{sec:denoising}

CER's significance is sharply sensitive to $K$, the number of positions
distinguishing the candidate from its reference peer. The joint per-read error
probability is $\prod_i q_{\text{ctx},i}$, so a spurious additional differing
position reduces $q_\text{ctx,joint}$ by roughly a factor of $q_\text{ctx}$
($\sim\!1/100$ at non-HP contexts) and inflates \texttt{cer\_factor}
accordingly. A noisy cluster whose consensus carries a handful of erroneous
positions will therefore compute a dramatically inflated factor against its
true peer, obscuring the distinction between real variants and cluster
artifacts.

This places a specific demand on the upstream pipeline: CER is a meaningful
discrimination metric only for clusters whose consensus is a faithful
representation of the reads that compose it. Before CER evaluation, speconsense
applies several denoising mechanisms whose purpose is to keep $K$ honest:

\begin{itemize}
\item \textbf{Outlier removal during final consensus generation.} MAD-based
detection of read-identity outliers excludes single-read noise that would
otherwise inflate consensus-level disagreement at isolated positions.

\item \textbf{Homopolymer-equivalence merging of near-identical post-phasing
clusters} (pre-merge step), collapsing splits whose only distinguishing
feature is HP length variation --- the dominant systematic error mode that
would inflate $K$ by one or more positions per HP run.

\item \textbf{MSA-based variant merging in summarize with IUPAC consensus},
further collapsing post-phasing clusters whose remaining differences are
dominated by positions already classified as ambiguity-compatible under the
specimen's error profile.

\item \textbf{A cluster-wide homogeneity statistic, \texttt{err\_factor}},
reported per cluster and filtered in summarize via \texttt{--max-err-factor}
(default 1.5). \texttt{err\_factor} is the ratio of observed per-column
disagreement to the $q_\text{ctx}$-predicted disagreement rate, pooled across
the cluster's MSA (non-gap consensus columns). Values near 1.0 indicate a
cluster consistent with per-position basecaller noise; values well above 1.0
indicate internal heterogeneity beyond what the platform's error model
predicts. Unlike \texttt{cer\_factor}, \texttt{err\_factor} is peer-independent:
it distinguishes novel-but-real sequences (low \texttt{err\_factor}, potentially
high \texttt{cer\_factor}) from noise aggregates (high \texttt{err\_factor},
often also spuriously high \texttt{cer\_factor}). Filtering on
\texttt{err\_factor} before CER evaluation removes clusters whose $K$
measurement would be unreliable regardless of the pairwise comparison. The
raw components (\texttt{err\_factor\_obs\_sum}, \texttt{err\_factor\_exp\_sum},
\texttt{err\_factor\_cols}) are persisted in the metadata JSON
(Section~\ref{sec:reproducibility}) so the factor can be recomputed under a
revised $q_\text{ctx}$ table.
\end{itemize}

None of these mechanisms replaces CER; they make CER's $K$ input trustworthy.
The natural place in the pipeline for this preprocessing is between phasing
and pairwise CER validation, which is where speconsense applies them.

\section{Validation}
\label{sec:validation}

\textbf{This section is a placeholder. Empirical content will be added as
Phase 1 output accumulates.}

Schema-2.0 metadata is now emitted by the current implementation. A full
ONT98 run (955 specimens) at \texttt{dorado-v5.0} is available as initial
input to the validation work below.

The planned validation reproduces the structure of the original CER
validation~\cite{cer_paper}~\S 8 but with the context-aware framework:

\begin{itemize}
\item Apply context-aware CER to the ONT98 dataset (955 specimens) and
compare per-variant decisions against the existing raw-$p^*$ implementation
at default settings.

\item Distribution of \texttt{cer\_factor} on the 955-specimen workload,
partitioned by variant class ($K{=}1$ non-HP, $K{=}1$ HP, $K{>}1$ mixed), to
empirically motivate the \texttt{--min-cer-factor} default
(Section~\ref{sec:factor}).

\item Apply to run116 (127 specimens, herbarium type material) and quantify
the impact of the cross-run $q_\text{ctx}$ scaling factor identified
in~\cite{hp_paper}~\S 5.4 on accept/reject decisions.

\item Sensitivity analysis: vary $q_\text{ctx}$ by $\pm 50\%$ at each context
class and quantify the fraction of decisions that flip. This characterizes the
framework's robustness to $q_\text{ctx}$ misspecification under realistic
uncertainty.

\item HP-variant rescue: identify HP length variants that the current pipeline
discards via blanket normalization but the context-aware framework would
accept. Manual review of a sample to assess biological plausibility.

\item Specimen-level outcomes: confirm that no specimens are lost (zero
organism dropouts), consistent with the original CER validation.
\end{itemize}

Expected results: a moderate increase in HP-variant calls (mostly $L = 1$--$3$
variants currently discarded), a modest reduction in marginal substitution
variants at HP-interface positions (where the empirical $q_\text{ctx}$ is
slightly lower than the current uniform assumption), and stable specimen-level
counts.

\section{Limitations}

The framework as presented inherits the limitations of the underlying CER
paper~\cite{cer_paper} and adds a few specific to the context-aware extension.

\paragraph{Cross-read error correlation.} The binomial null assumes
independence of errors across reads. Reads sharing the same PCR template can
exhibit correlated errors (PCR-introduced substitutions, amplification bias);
these are physically present in the library and not addressed by the CER
framework, which establishes physical library presence rather than biological
reality.

\paragraph{Within-read non-independence at HP positions.} HP length errors are
signal-level events affecting the entire run at once, not independent per-base
errors~\cite{hp_paper}~\S 3.4. The per-position $q_\text{ctx}$ tabulated in
the empirical analysis is a compact parameterization for cross-context
comparison; the binomial model used here is correct because it operates
across reads (independent draws from the sequencing process), not within a
read. Practitioners should not interpret the per-position $q_\text{ctx}$ for
HP positions as a mechanistic rate.

\paragraph{Non-stationary error rates within a run.} If error rates drift
during a sequencing run (flow cell degradation, temperature variation), the
binomial model's identical-trials assumption is violated. Phase 2 per-run
estimation partially mitigates this for between-run drift but cannot address
within-run drift.

\paragraph{$q_\text{ctx}$ provenance for novel basecaller versions.} Shipped
$q_\text{ctx}$ tables are derived from empirical analysis of specific
basecaller versions~\cite{hp_paper}. New basecaller releases require either
re-characterization or use of the closest available table with explicit
acknowledgment of the mismatch. Per-run estimation (Phase 2) addresses this
but is not yet implemented.

\paragraph{Long HP runs ($L \geq 6$).} The framework does not extend to
$L \geq 6$ HP variants; these continue to be blanket-discarded as in the
existing pipeline. The HP error analysis~\cite{hp_paper}~\S 8.1 documents that
accuracy degrades sharply at $L \geq 6$ and variant calling becomes
unreliable; this is a property of the data, not the framework.

\paragraph{Biological variant priors.} The framework treats all positions as
having the same prior probability of carrying a real biological variant. In
ITS, conserved regions (5.8S) and hypervariable regions (ITS1, ITS2) have
very different prior variant densities. A Bayesian extension could incorporate
position-specific priors; the current framework does not.

\section{Conclusion}

The CER framework~\cite{cer_paper} provides a transparent statistic for variant
significance that is interpretable against platform error knowledge. Its
extension to context-dependent error rates --- demanded by the empirical
characterization of HP errors~\cite{hp_paper} --- requires changes to the
inputs of the CER equation, not to its structure. The unified equation
(Section~\ref{sec:unified}) accommodates per-context error rates and mixed-
context multi-position variants without abandoning the binomial framework.

The CER factor (Section~\ref{sec:factor}) is the practical metric that recovers
cross-variant comparability under context-aware $q_\text{ctx}$. It preserves
the human-interpretable units that motivated the original framework, degrades
linearly under realistic $q_\text{ctx}$ uncertainty, and re-scales cleanly
when the empirical error table is updated.

The implementation in speconsense (Section~\ref{sec:implementation}) integrates
the framework with existing pipeline structure: the $N$ denominator follows
the conservative choice at each stage, identity groups define the pairwise
comparison set, and HP variant detection is reordered so that $L \leq 5$ HP
length differences become candidate variants subject to CER rather than being
unconditionally discarded.

The metadata reproducibility design (Section~\ref{sec:reproducibility}) ensures
that stored output can be reinterpreted under updated assumptions. As
basecaller chemistry evolves and empirical error tables improve, no
recomputation from raw reads is required --- the pipeline's conclusions remain
adaptable to the moving target of nanopore error characterization.

Phase 2 work --- per-run $q_\text{ctx}$ estimation, an empirical-distribution
model for HP variants that captures within-read signal-level dependencies, and
incorporation of biological-variant priors --- remains future work, scoped
beyond the initial context-aware deployment.

\subsection*{Acknowledgments}

Substantial portions of the text were drafted with the assistance of Claude
(Anthropic, Opus 4.7). The core ideas, framework design, and implementation
choices were provided by the author; the AI contributed to exposition,
\LaTeX\ formatting, and iterative drafting under human direction. The author
takes full responsibility for the content.

\appendix

\section{Metadata JSON schema (full example)}
\label{app:schema}

A representative example of the schema-2.0 metadata emitted by the current
implementation (condensed for exposition; the \texttt{parameters} block carries
the full set of CLI arguments in practice). Two identity groups are shown: a
populated group (\texttt{g1}) with an anchor and two non-anchor variants, and
a singleton group (\texttt{g2}).

\begin{lstlisting}
{
  "schema_version": "2.0",
  "version": "0.7.7",
  "sample_name": "ONT08.75-C10--iNat262724197",
  "timestamp": "2026-04-20T18:09:14.624236",

  "parameters": {
    "algorithm": "graph",
    "min_identity": 0.9,
    "min_size": 3,
    "max_sample_size": 100,
    "presample_size": 1000,
    "min_variant_frequency": 0.1,
    "min_variant_count": 3,
    "significance_level": 1e-05,
    "group_identity": 0.85,
    "error_model": "dorado-v5.0"
  },

  "input_file": "ONT08.75-C10--iNat262724197.fastq",
  "total_input_reads": 1000,

  "identity_groups": [
    {
      "group_id": "g1",
      "group_N": 974,
      "members": ["1.v1", "1.v2", "1.v3"]
    },
    {
      "group_id": "g2",
      "group_N": 7,
      "members": ["2.v1"]
    }
  ],

  "variants": [
    {
      "cluster_id": "1.v1",
      "identity_group": "g1",
      "group_rank": 1, "variant_rank": 1,
      "M": 462, "N": 974,
      "K": null, "n_sites": null, "context_tags": null,
      "q_ctx_per_position": null,
      "compared_against_idx": null,
      "cer_factor": null, "cer_pstar": null,
      "err_factor": 0.303,
      "err_factor_obs_sum": 3.13,
      "err_factor_exp_sum": 10.33,
      "err_factor_cols": 781
    },
    {
      "cluster_id": "1.v2",
      "identity_group": "g1",
      "group_rank": 1, "variant_rank": 2,
      "M": 133, "N": 974, "K": 2, "n_sites": 560,
      "context_tags": ["non-hp-sub", "non-hp-sub"],
      "q_ctx_per_position": [0.0059, 0.0059],
      "compared_against_idx": 0,
      "cer_factor": 46.500, "cer_pstar": 0.274,
      "err_factor": 0.445,
      "err_factor_obs_sum": 4.61,
      "err_factor_exp_sum": 10.35,
      "err_factor_cols": 781
    },
    {
      "cluster_id": "1.v3",
      "identity_group": "g1",
      "group_rank": 1, "variant_rank": 3,
      "M": 49, "N": 974, "K": 1, "n_sites": 561,
      "context_tags": ["hp-l3-A-del1"],
      "q_ctx_per_position": [0.0097],
      "compared_against_idx": 0,
      "cer_factor": 2.085, "cer_pstar": 0.0202,
      "err_factor": 0.453,
      "err_factor_obs_sum": 4.67,
      "err_factor_exp_sum": 10.32,
      "err_factor_cols": 780
    }
  ]
}
\end{lstlisting}

\section{Default $q_\text{ctx}$ table for Dorado v5.0}
\label{app:qctx_table}

Drawn from~\cite{hp_paper} Tables 4 and 5. Values are filtered per-position
error rates (after biological-variant filtering) for the ont98 dataset
(R10.4.1, Dorado SUP v5.0.0). Per-base and per-direction sub-dimensions are
pooled for the reasons given in Section~\ref{sec:qctx_pooling}: the effects
are smaller than the run-to-run noise floor and cross-cell counts are too
sparse to resolve the finer segmentation stably.

\begin{table}[htbp]
\centering
\caption{Shipped $q_\text{ctx}$ values for Dorado v5.0 (per-position error
rate). This is the complete contents of \texttt{dorado-v5.0.yaml}; per-base
and per-direction variation is folded into each entry.}
\label{tab:default_qctx}
\small
\begin{tabular}{lcl}
\toprule
Context tag & $q_\text{ctx}$ & Source \\
\midrule
\texttt{non-hp-sub}            & 0.0059 & \cite{hp_paper} Table 5 (ont98 substitution) \\
\texttt{non-hp-indel}          & 0.0108 & \cite{hp_paper} Table 5 (deletion + insertion) \\
\midrule
\texttt{hp-l1}                 & 0.0067 & \cite{hp_paper} Table 4 \\
\texttt{hp-l2}                 & 0.0083 & \cite{hp_paper} Table 4 \\
\texttt{hp-l3}                 & 0.0097 & \cite{hp_paper} Table 4 \\
\texttt{hp-l4}                 & 0.0099 & \cite{hp_paper} Table 4 \\
\texttt{hp-l5}                 & 0.0113 & \cite{hp_paper} Table 4 \\
\midrule
\texttt{hp-l6+}                & N/A    & blanket discard, \cite{hp_paper}~\S 8.1 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Future refinement.} Finer segmentation --- per-base (e.g.,
\texttt{hp-l5-G = 0.0140} per~\cite{hp_paper}~Table~10) or per-direction
(\texttt{hp-l3-A-del1} vs \texttt{hp-l3-A-ins1}) --- is straightforward to
add by extending the YAML with additional keys. The lookup function already
accepts any string key, and the context tag persisted in the metadata JSON
records the full $(b, L, \delta)$ triple. Downstream recomputation under a
finer table therefore requires no upstream re-run. The current simplification
reflects the data available at publication time: with run-to-run variation
of $\sim\!0.6\times$~\cite{hp_paper}~\S 5.4, the finer cells would introduce
false precision rather than discrimination.

HP-interface substitutions inherit the \texttt{non-hp-sub} rate automatically:
the context classifier reports them as \texttt{NON\_HP\_SUB} tags per the
finding in~\cite{hp_paper}~\S 4 that interface substitution rates are
indistinguishable from non-HP. No separate table entry is needed.

\begin{thebibliography}{99}

\bibitem[Walker, 2026a]{cer_paper}
J.~Walker.
\newblock Critical error rate analysis for {MSA}-based variant phasing in
nanopore amplicon sequencing.
\newblock Technical report, 2026.

\bibitem[Walker, 2026b]{hp_paper}
J.~Walker.
\newblock Homopolymer error rate analysis for nanopore amplicon sequencing.
\newblock Technical report, 2026.

\bibitem[Walker, 2024]{speconsense}
J.~Walker.
\newblock Speconsense: High-quality clustering and consensus generation for
{Oxford Nanopore} amplicon reads.
\newblock \url{https://github.com/joshuaowalker/speconsense}, 2024.

\bibitem[Delahaye and Nicolas, 2021]{delahaye2021}
C.~Delahaye and J.~Nicolas.
\newblock Sequencing {DNA} with nanopores: Troubles and biases.
\newblock \emph{PLoS ONE}, 16(10):e0257521, 2021.

\bibitem[Sereika et~al., 2022]{sereika2022}
M.~Sereika, R.~H. Kirkegaard, S.~M. Karst, T.~Y. Michaelsen, E.~A.
S{\o}rensen, R.~D. Wollenberg, and M.~Albertsen.
\newblock {Oxford Nanopore R10.4} long-read sequencing enables the generation of
near-finished bacterial genomes from pure cultures and metagenomes without
short-read or reference polishing.
\newblock \emph{Nature Methods}, 19:482--488, 2022.

\bibitem[Wick et~al., 2023]{wick2023}
R.~R. Wick, L.~M. Judd, and K.~E. Holt.
\newblock Assembling the perfect bacterial genome using {Oxford Nanopore} and
{Illumina} sequencing.
\newblock \emph{PLOS Computational Biology}, 19(3):e1010905, 2023.

\bibitem[Lee, 2003]{lee2003}
C.~Lee.
\newblock Generating consensus sequences from partial order multiple sequence
alignment graphs.
\newblock \emph{Bioinformatics}, 19(8):999--1008, 2003.

\bibitem[Vaser et~al., 2017]{vaser2017}
R.~Vaser, I.~Sovi\'{c}, N.~Nagarajan, and M.~\v{S}iki\'{c}.
\newblock Fast and accurate de novo genome assembly from long uncorrected reads.
\newblock \emph{Genome Research}, 27(5):737--746, 2017.

\end{thebibliography}

\end{document}
